---
title: "Report"
author: "H.S. Reefman"
date: "2-11-2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(ggplot2)
library(tidyr)
library(gridExtra)
library(plyr); library(dplyr)
library(grid)
library(patchwork)
```

# Introduction

```{r dataset}
load("~/Bio-informatica/BFV3/Semester1/Themaopdracht09/tedsd_puf_2019.RData")

p <- "CASEID|AGE|EDUC|^EMPLOY|ALCDRUG|^SUB|^FREQ[1|2|3]|^ROUTE|^FRSTUSE|LOS|NOPRIOR|PSYPROB|DSMCRIT|REASON"

# get relevant data from data set
data <- df[, grep(pattern=p, colnames(df))]

```

The research topic of this project is the treatment episodes of patients. 
The dataset used in this project is from the Treatment Episode Data Set (TEDS), 
an American data system of annual discharges from substance use treatment facilities.
The used dataset is a TEDS-D dataset, which means it includes discharges from substance
use treatment facilities. TEDS-D contains records on admissions of people of 12-years and older,
and includes information of admission, substance use characteristics and discharges.
The dataset includes information from patients who where at the facility in 2019.
For this project some personal and substance use characteristics information are collected from the dataset. 
With this dataset a model can be made. In this project is researched whether the length of treatment can be predicted. 
The model is made by the following question: "Can the length of treatment be predicted 
based on the substance use, frequency of use and the age at first use by using machine learning?"
To answer this question information is collected from the dataset. The personal information
about the patients that is used is age.
Other information is about substance characteristics, for example; substance use type, 
substance use at admission and discharge and route of administration.
The information about treatment episodes are also covered, this includes length
of stay in treatment.

# Materials and Methods

Exploratory data analysis and data preparation is done with the R programming language using Rstudio.
Plots were made making use of the 'ggplot2' packages. For the layout of the plots the packages 'grid' and 'gridExtra' were used. Other packages that were used include 'tidyverse', 'tidyr', '(d)plyr', with as function to format tables. 
The model was built with Weka (version 3.8.5), a free data mining software written in Java.
To built the model the clean dataset was inserted into the Weka Explorer and the performance of all standard machine learning algorithms were investigated. The machine learning algorithms investigated are:
 
- ZeroR
- OneR
- Naïve Bayes
- Simple Logistic
- Nearest Neighbor (IBK)
- J48

The classifications were carried out using 10-fold cross validations and the relevant quality metrics were recorded. These include: speed, accuracy, true positives (TP), false positives (TP), true negatives (TN) and the false negatives (FN). This is done in the Weka Experimenter.  

Statistical tests were applied making use of the Weka Experimenter again. And some Meta learners were investigated. The findings were recorded and discussed. 

In the end a ROC curve visualization and a learning curve were created. 

Because the dataset included so many observations (around the 1.7 milion) a sample of the dataset
was taken. To make sure this sample was a representative sample of the dataset, first the full dataset was investigated with the machine learning ZeroR. Also the sample was investigated with ZeroR (see Results). 

The logbook and data of this research can be found in a repository on Github.com:
https://github.com/Susanreefman/Themaopdracht09

At last a command-line application (wrapper) was made in the Java programming language. 
Code for the wrapper can be found in a repository on Github.com:
https://github.com/Susanreefman/JavaWrapperThema09



# Results

## EDA
In figure 2 multiple stacked barplots are shown. 
The top barplot depicts the age of first use of the patients. The age is divided in multiple groups, in which the first groups is 11 years of younger and the last group is 30 years of older. Every bar is divided into 3 classes. The pink class is the age of first using the primary substance of the patients. The blue class is the age of first using the secondary substance, and the green class the tertiary substance. The results seem normally distributed with a large amount of patients who started between 15 and 17 years old with using their primary, secondary or tertiary substance.

The second barplot shows the frequency of use of patients. Unlike the groups the frequency of use in which the data is divided, the 'NA' group is extremely large. Out of proportion is are the missing values of the frequency of the tertiary substance used. This could be because patients do not have a tertiary substance that is used. The frequency for the secondary and tertiary substance is lower, most patients just use one substance. The most patients use their primary substance daily. 

The third barplot depicts the usual route of administration of patients. 
Most patients use the oral route of administration for their primary substance use.
Most patients that have a secondary substance use, use the smoking route of administration
of that substance. Again large amount of patients have missing information for the route of the secondary and tertiary substance, these are the patients that said they did not have a secondary or tertiary substance they use. 

Figure 3 shows the number of patients and their substance use. 
Not surprisingly is the blue (tertiary substance) very high for the None of substance of use. Not all patients have multiple substances they are using. However, 
there are a few patients that filled in that the have none as primary substance of use (pink). 
Alcohol, Heroin and Methamphetamine are the most primary substances of patients. Cocaine and Marijuana or Hashish are mostly the secondary substance used. 

The boxplot in figure 4 shows the distribution of patients length of stay in treatment in days
versus the substance use type which they get treated for. The mean of length of stay 
when they indicated "none" as substance use type is 9 days. For alcohol and other drugs
the treatment is much longer. For alcohol as substance use type the average length of stay is 22 days.
For the patients with a substance use type of only drugs or alcohol and other drugs the average length
of stay is 28 days. The length of stay has a large difference in all substance use type groups. 


## Cleaning dataset

Because the dataset is very large and most of the data is not included in this project.
Only the characteristics of the use of patients is necessary, which include: Type of use (alcohol, drugs or both), the substance of use, the frequency of use, the route of inhalation and the age of first use. The length of stay is also necessary in this project. 
Because the dataset is this large, the patients which have information that is not available is removed from the dataset. This includes around 200.000 patients, the dataset still includes more than 1.5 million patients. Removing the missing values makes it also easier for the machine learning algorithm to classify instances. 
Thereafter a sample of the dataset is taken. Both the dataset and the sample are written to csv files, to 
be uploaded to Weka. 

## WEKA

The first action that was taken in Weka is validating the sample, both datasets were
investigated with the ZeroR machine learning algorithm. In the table beneath, can be seen
that the percentage of instances (in)correctly classified are not significant different.
Taking the sample as the dataset to investigate further.
The performance of standard machine learning algorithms was investigated. The classifier and
attribute investigated were noted as well as the percentages (in)correctly classified instances. 

```{r fig.cap="table 1:"}
attributes <- c("type drug", "sub", "freq", "frstuse", "los")
data_zeroR_correct <- c("%53.9", "33.7%", "45.5%", "24.4%", "13.6%")
data_zeroR_incorrect <- c("46.1%", "67.3%", "54.5%", "75.6%", "86.4%")
sample_zeroR_correct <- c("%53.8", "33.6%", "44.5%", "24.5%", "13.8%")
sample_zeroR_incorrect <- c("46.2%", "67.4%", "56.5%", "75.5%", "86.2%")

knitr::kable(data.frame(attributes, data_zeroR_correct, data_zeroR_incorrect, sample_zeroR_correct, sample_zeroR_incorrect))

```


```{r fig.cap="table 2: "}
weka <- read.table("~/Bio-informatica/BFV3/Semester1/Themaopdracht09/Weka_results/weka.csv", 
                  sep = ',',
                  header = T)

knitr::kable(weka)
```


## Machine learning

```{r}
c <- c("ZeroR", "OneR", "Naive Bayes", "Ikb", "SimpleLogistics", "SMO")
d <- c("13.6%", "10.1%", "85.7%", "17.1%", "73.0%", "11.5%")
e <- c("86.4%", "89.9%", "14.3%", "82.9%", "27.0%", "88.5%")

frame <- data.frame(d,e)
row.names(frame) <- c
colnames(frame) <- c("Correctly classified", "Incorrectly classified")

frame
```

Waarom Naive Bayes en SL:
Meeste goede, duurt niet lang

Optimalizatie van NB en SL:
  algorithm
  
  meta learners


For further optimalization Naive Bayes and SimpleLogistics algorithms are choosen. 
First Naive Bayes is investigated.
Changing to the `Train/Test Percentage Split (data randomized)` test option:
With 10 runs and the split percentage ranging from 40 to 90%, the algorithm
classifies between the 70.66% (40% split percentage) and 70.74% (66% split percentage).
This is not better than the original Naive Bayes algorithm. 
Thereafter the k-fold cross-validation was tested. But also here, no optimalization 
by changing the number of folds higher or lower.

Moving on to the SimpleLogistics algorithm:
Changing again the `Train/Test Percentage Split (data randomized)` test option:
With 10 runs and the split percentage ranging from 40 to 90% the algorithm, the
algorithm classified every time 72% of the instances correctly. This is not worse, 
but also no optimalization. 
Optimalization was achieved with increasing the number of folds with the k-fold cross-validation
test-option. With 15-folds, 75% of the instances were correctly but the time performing
the algorithm doubled, so in general would the increase not be sufficient.


naive bayes
Train/Test Percentage Split (data randomized) training% 40.0, number of runs = 10 --> 70.66
Train/Test Percentage Split (data randomized) 66.0, number of runs = 10 --> 70.74
Train/Test Percentage Split (data randomized) 80.0, number of runs = 10 --> 70.71
Train/Test Percentage Split (data randomized) 90.0, number of runs = 10 --> 70.68
Niet beter dus laten bij cross-validation
5-fold crossvalidation -> 78.70

sl
Train/Test Percentage Split (data randomized) training% 40.0, number of runs = 10 --> 72.41
Train/Test Percentage Split (data randomized) 66.0, number of runs = 10 --> 72.41
Train/Test Percentage Split (data randomized) 80.0, number of runs = 10 --> 72.40
Train/Test Percentage Split (data randomized) 90.0, number of runs = 10 --> 72.40


# Conclusion and Discussion

Dataset bestond uit heel veel waarden dus miss andere column meer invloed op length of stay.


<!-- Overall, as can be seen in the figures (see Results), the data is widely spread.
The age of patients and the age of first use of patients seem normally distributed. 
The attributes of variables can be discussed. In some cases the attributes are vague
and can be wide interpreted. This is in the case of frequency of use, there 
are three groups in which the data is divided: No use in the past month, Some use 
and Daily use. As the first group is very clear, the second group is to vague. 
What is meant by some use? Also the third group, daily use, can be up for discussion;
how much is the substance used daily. --> 


## Project proposal for minor
